---
output: html_document
---
Activity Tracker Data Analysis
========================================================

## Reading the data

Activity tracker data is interesting! Let's load and analyze the dataset provided for this assignment, which is stored locally as *"activity.csv"*. We'll use the *read.csv()* function to read in the data and store it in the cache so it doesn't need to be repeatedly reloaded. And while we're at it, let's make a summary of the data to see what's there.

```{r cache = TRUE}
activity_data <- read.csv("./activity.csv")
summary(activity_data)
```

Alright, so based on our summary we see that there are a fair number of "NA" values in our *steps* variable; this might come up a fair amount going forward. The data appear to be in a pretty usable format, so having noted that, let's move on to some more analyses. 

## Steps Per Day

A natural starting point in this dataset is to ask how many steps were taken each day. Let's create a histogram to show this. The first thing we need to do is sum each day's steps so that we have one value per day. We can use *tapply()* to accomplish this, and then we'll plot the resulting totals using *hist()*.

```{r}
totals <- tapply(activity_data$steps,activity_data$date,sum,na.rm=TRUE)
hist(totals,breaks = 20,col="steelblue",xlab = "Total Steps",
     main = "Total Steps per Day")
```

At first glance, this looks a little odd in that there are 10 days with only 0-1000 steps; however, upon looking at the raw dataset, this seems to be correct. For instance, the first day ("2012-10-01") does not have any steps recorded and the second day ("2012-10-02") has only a couple hundred. So even though this looks strange at first, it certainly seems to correspond to the data. 

We've looked at the distribution of steps, but how about the center of the distribution? We can calculate the mean and median straightforwardly by using the appropriate functions, *mean()* and *median()*.  

```{r}
mean(totals)
median(totals)
```

We see based on these outputs that both values are around 10,000. To better understand them, let's plot them on our histogram

```{r}
totals <- tapply(activity_data$steps,activity_data$date,sum,na.rm=TRUE)
hist(totals,breaks = 20,col="steelblue",xlab = "Total Steps",
     main = "Total Steps per Day w/Mean and Median")
abline(v = mean(totals),col = 'green',lwd = 2)
abline(v = median(totals),col = 'red',lwd = 2)
legend("topright",col = c('green','red'), legend = c('Mean','Median'),
       lty = 1, lwd = 2)
```

As we expected, these two lines both appear in the vicinity of 10,000 steps. Now that we've characterized the steps for each day, let's take a look at this individual's average day and see what we can say about them.

## Average Daily Activity

Now instead of looking at day to day steps, let's see what happens if we consider our "average" day. The average day is defined by 5 minute time intervals, ranging from 0 to 2355 as shown by our previous summary. So let's use tapply to find the mean of the steps in each interval.

```{r}
daily <- tapply(activity_data$steps,activity_data$interval,mean,na.rm=TRUE)
```

We can now use these mean values along with their labels to plot a time series plot of steps against time interval.
```{r}
plot(d)
